Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-15 tex/paper.tex
draft

@@

\begin{abstract}
    This study investigates the impact of the Consumer Price Index, Bank of England Base Rate, US Dollar (USD) and British Pound Sterling (GBP) exchange rate and the M3 money supply measure on the FTSE 250 index, and assesses the effectiveness of radial basis function neural networks as a tool to forecast the price of the FTSE 250 index based on the same set of macroeconomic indicators. This study shows quantitative estimates for the long-term and short-term impacts of the aforementioned macroeconomic indicators on the price of the FTSE 250 index, and demonstrates the effectiveness of radial basis function neural networks as a short-term forecasting tool for the price of the FTSE 250 index, thus providing an alternative to econometric forecasting models.

@@

The stock market has played a central role in the economic and political ascent of many nations ever since its inception [1].  However, whenever stock market instabilities have arisen in history they have had far-reaching destabilising effects. The Mississippi Bubble was an early example of the inherent dangers within stock markets in the form of speculative bubbles [2]. In this particular case, the stock market crashed as result of an unsustainable rise in share prices fueled by speculation, leading to panic selling, a collapse in confidence, and a massive financial meltdown that rippled through the entire French economy.. Another example is the bursting of the dot-com bubble which had a profound impact on the stock market, leading to a significant crash, particularly in technology stocks. It triggered a severe stock market crash by leading to a rapid sell-off of overvalued tech stocks, which then spread to the broader market. The collapse of key companies, along with forced selling due to margin calls, and the widespread loss of investor confidence created a downward spiral that devastated the stock market.

Nevertheless, stock markets provide essential benefits by allowing companies to raise capital, offering investment opportunities for individuals, and supporting economic growth through efficient capital allocation. They enhance corporate governance through transparency and accountability, facilitate price discovery and market liquidity, and contribute to government financing through taxes.(Shahbaz et al., 2016).

The stock market is highly sensitive to macroeconomic indicators because these indicators provide insights into the overall health and direction of the economy. Stock markets react to macroeconomic indicators because these signals help investors anticipate future economic conditions, corporate profits, and investment returns. Positive indicators tend to boost stock prices, while negative indicators can lead to declines in stock prices.

@@

Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability \citep{boe199section}. 
The empirical research supports this theory. \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003.
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange.
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the FTSE100. 
In addition, the flight-to-quality phenomenon that igher interest rates also encourage a shift from stocks to fixed-income 
assets such as bonds. \citeboth{asgharian2016} provided evidence for this in developed stock markets, including the UK. This suggests interest rates 
negative impact the stock market.

@@

One of the foundational studies on the relationship between the money supply and 
stock market prices was \citeboth{palmer1970}, who analysed the impact of M1 money supply on the S$\&$P 425 Industrial Stock Index. 
He posited that trends in the nation's money stock could lead the 
private sector to adjust wealth portfolios, resulting in 
predictable movements in stock prices. \citeboth{homa1971} further argued that 
stock prices are significantly impacted by the risk-free interest rate, 
which is a function of money supply. This relationship was disputed by \citeboth{pesando1974}, 
who critiqued the theoretical and empirical methodology used in their study. In addition, \citeboth{sorensen1982} and 
\citeboth{bernanke2005} argue that only the unanticipated component of money supply changes 
affects the stock market price. Later research, however, such as \citeboth{bahloul2017}, 
\citeboth{synek2024}, and \citeboth{pícha2017} has found that the price of developed stock markets, such as the US and UK's, is in fact significantly influenced by the money supply. 
The nature of this relationship is disputed and varies across countries. \citeboth{homa1971} established a positive relationship between the 
stock market and the money supply. This was disputed by \citeboth{sellin2001}, who presents a negative theoretical relationship betwen the stock market and money supply. \citeboth{olawale2014} empirically found a significant positive impact by the money supply on stock returns in the US, but a negative one in the case of the UK.

@@

During the data collection process, it was necessary to deal with an issue raised by the fact that the Bank of England’s Monetary Policy Committee (MPC) 
does not convene on a monthly basis, and are not obligated to 
announce a rate on a regular basis. To address the resulting irregularities the data was adjusted using the following algorithm:

@@

The CPI, with an average of 88.36, has 
experienced significant fluctuations, highlighting periods of high inflationary pressure such as in the late 2000s 
near the global financial crisis, and during the onset of the Russia-Ukraine conflict in 2022, and periods of stable, low
inflation such as the early to late 2010s.

@@

Before constructing the ARDL model a natural logarithmic transform is applied
on the variables, this is done because in a log-transformed model,
the coefficients can be interpreted as elasticities (percentage changes). Having done this,
the level at which the variables are stationary is checked.
Various techniques exist for stationarity testing, however in this study the 
Augmented Dickey-Fuller (ADF) unit root test was used. The optimum lag for the variables in the ARDL model was then found using the Schwarz
Information Criterion.

The implicit form \eqref{eq:implicit} is represented in the ARDL model as:

@@

Here, the intercept term, represents the baseline level of \(\Delta \text{FTSE}_{250_t}\), is given by $\alpha_0$; the 
remaining $\alpha_i$ terms represent the short-run relationships 
between the variables, and $\phi_j$ represent the long-run relationships.

@@

for, $i,j = 1,2,3,4,5$, with $i\neq j$. The F-statistic is computed for the joint
significance of the lagged variables and compared against the critical values
given by \citeboth{pesaran2001} to determine whether cointegration exists, 
and at what significance level. If cointegration exists, then 
the long-run coefficients $\kappa_i$ are estimated to find the cointegration equation:

@@

Then, an Error Correction Model (ECM) is constructed:

@@

where the error correction term ECT is found by calculating the residual 
produced by \eqref{eq:two}:

@@

The parameter $\gamma$ describes the speed at which short-run deviations from the long-run equilibrium
are adjusted. A negative $\gamma$ implies that there exists a correction mechanism that converges the deviation 
to the long-run equilibrium. To estimate the short-run coefficients, $\alpha_i$ and $\beta_k$, as 
well as the short-run correction speed $\gamma$ OLS regression is used.

@@

First, a natural logarithmic transform was applied on the variables, and 
then conducting the ADF unit root test on the data. The null hypothesis of the 
ADF Unit Root test is that of non-stationarity, and the alternate hypothesis 
is that the variables are stationary.

@@

Subsequently, the optimum lag of the variables was calculated using the 
Schwarz Information Criterion, to obtain the optimum model 
of ARDL(5, 4, 7, 7, 4). The resulting F-statistic of $4.0798$ was compared against the critical values
given by \citeboth{pesaran2001}, providing evidence of cointegration 
at the $10\%$, $5\%$ and $2.5\%$ significance levels, but not at the $1\%$ 
level.

@@

The long-run and short-run coefficients were then estimated, and presented 
alonsgide the diagnostic test results which confirm that the model does not
display autocorrelation, heteroskedascity, non-normality and the cumulative 
sum of residuals and squared residuals are both stable, indicating that the 
relationship between the variables has not changed over time. The error
correction term is also given, and it is negative, as expected.

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & $-$ & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{***}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{****}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.1001^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

Unsurprisingly, the USD/GBP exchange rate is significantly related to the FTSE250's movements. It 
should not be forgotten that, although it is considered to be a more domestic indicator of the LSE, 
at times $50\%$ of the FTSE250's earnings have come from abroad, indicating that 
the value of the Pound has a significant impact on this stock market. 
In the long run, the appreciation of the Pound would increase the relative 
cost of UK exports increases, thereby reducing the 
competitiveness of FTSE250 companies in international markets. This in turn would lower their earnings
and depress the FTSE250. Moreover, foreign investors may view the stronger Pound as a signal of reduced returns on their investments in the UK, which can lead to decreased investment in the FTSE250. 
In the short-term, however, before structural shifts can materialise we see that the
weakening of the Pound has a positive effect. As the USD has been 
the global reserve currency for many decades, the depreciation of the Pound 
against it will lead to increased revenue in Pound terms for FTSE250 companies with substantial overseas operations. This is because their foreign earnings, when converted back into Pounds, will be worth more, potentially increasing profits. The findings 
are in line with \citeboth{neifar2023} and \citeboth{khan2018}.

@@

The error correction term 
is negative and statistically significant ensuring that long-run equilibrium 
can be achieved. The adjustment speed of $0.1001$ indicates that 
roughly $10\%$ of the fluctuation in the previous month's price caused by the 
macroeconomic variable movements is adjusted back to the long-run equilibrium in the current month.
However, only increases in the interest rate and money supply are able to 
significantly contribute towards this correction, with $1\%$ increases in the former 
leading to a $-0.03\%$ change in the FTSE250 price 7 months later, and in the case of the latter
a $-0.53\%$ change after a 4-month lag.

@@

\begin{figure}[h]
    \centering
    \hspace{-2.6cm} 
    \begin{tikzpicture}[node distance=3cm,
                        thick, 
                        main node/.style={circle, draw, font=\sffamily\Large\bfseries}]

      \node[main node] (1) {$y_{1}$};
      \node[main node] (2) [right of=1] {$y_{c}$};
      \node[main node] (3) [below left of=1] {$\phi_{1}$};
      \node[main node] (4) [below right of=2] {$\phi_{M}$};
      \node[main node] (5) [below right of=3] {$x_{1}$};
      \node[main node] (6) [below left of=4] {$x_{d}$};
      \node[main node] (7) [left of=3] {$\phi_{0}$};

      \draw[thick, dashed] (1) -- (2);
      \draw[thick, dashed] (3) -- (4);
      \draw[thick, dashed] (5) -- (6);
      \draw[thick] (1) -- (3);
      \draw[thick] (1) -- (4);
      \draw[thick] (2) -- (4);
      \draw[thick] (7) -- (1);
      \draw[thick] (7) -- (2);
      \draw[thick] (3) -- (1);
      \draw[thick] (3) -- (2);
      \draw[thick] (3) -- (6);
      \draw[thick] (3) -- (5);
      \draw[thick] (4) -- (5);
      \draw[thick] (4) -- (6);

    \end{tikzpicture}
    \caption{RBF-NN Architecture}
    \label{fig:your_label}
\end{figure}

@@

The radial basis function approach 
introduces a set of $N$ basis functions, one for each 
data point. The radial basis function, $\phi(\cdot)$ is some non-linear function of the Euclidean distance
$||\boldsymbol{\mathrm{x}} - \boldsymbol{\mathrm{x_n}}||$. The output of the mapping is then expressed as a linear combination of these basis functions:

@@

By introducing several modifications to the exact interpolation 
procedure, the RBF-NN
model is obtained, as outlined originally in \citeboth{broomhead1988}. 
The RBF-NN mapping takes the following form:

@@

where $\boldsymbol{c_j}$ is a RBF centre and $f$ is the radial basis function.

@@

The RBF-NN was utilised for both short-term and long-term 
predictions. For short-term forecasting, 
the model was tested over a 5-month horizon, based on the optimum lag calculated earlier, while the long-term predictions spanned 57 months. 
The data was differenced based on the recommendations of the 
unit root tests conducted earlier to achieve stationarity.

The data was normalised using a z-score normalisation. This is because
neural networks work best with data
that is between a specified range; it ensures that the data is roughly uniformly
distributed between the network inputs and outputs 
(Mendelsohn, 1993, Preprocessing Data for Neural Networks). Following this 
it was necessary to program the RBF-NN into Python. As no specific RBF package exists
in Python, the kernel regression approach outlined by 
\citeboth{bishop1995}, using the GaussianProcessRegressor package in Python.

@@

The Scikit GaussianProcessRegressor is based on the 
standard linear regression 
model, equipped with Gaussian noise,

@@

where $y$ differs from the function values $f(x)$ by noise which is assumed
to follow an independent and identically distributed Gaussian distribution
$\mathcal{N}_1$
\begin{align*}
    \varepsilon \sim \mathcal{N}_1 \bigl(0, \sigma_{n}^2\bigr)
\end{align*}
This assumption together with the model produces the 
probability density of the observations given the inputs, 
which is factored over cases in the training set, 
permissible because of the independence assumption,

@@

As this is a Bayesian model a \textit{prior distribution} over the parameters must be specified, 
expressing our beliefs about the prior
parameters before the observations are used. A covariance function or kernel
is specified according to the characteristics of the data; in our model this 
is the generalised inverse multiquadric

\begin{align}
    k(x,x') = \biggl( ||x-x'|| + \sigma^2\biggr)^{-\beta} \label{eq: gimq}
\end{align}
and used to construct the covariance matrix $K$, which along with 
a mean of $0$ forms the distribution of the the weights 
$$ w \sim \mathcal{N}\bigl(m(x), K\bigr)$$

The mean function $m(x)$ acts as a bias, providing a baseline expectation of the function values. Then, by Bayes' Theorem,

\begin{align}
    p(w|y, X) = \frac{p(y|X,w) p(w)}{p(y|X)}
\end{align}
Using this, the posterior distribution is then obtained

\begin{align}
    p(w|y, X) \sim \mathcal{N}\bigl(\frac{A^{-1}Xy}{\sigma_{n}^{2}}, A^{-1}\bigr) \label{eq:nine}
\end{align}
Tthe covariance matrix $A^{-1} = K^{-1} + \sigma_{n}^{-2}(XX^T$).
To make predictions on a test case all possible $w$, weighted by their probability, are averaged to obtain the
predicted output \citep{rasmussen2006}.

@@

First, the optimal hyperparameters in the covariance function \eqref{eq: gimq} 
are computed using GridSearch Cross Validation; these hyperparameters are:
\begin{enumerate}
    \item $\alpha$: a regularisation parameter which represents a small quantity
    that is added to the diagonals of the covariance matrix to ensure it is 
    invertible. 
    \item $\beta$: the power in the covariance function, which controls the shape of the function.
    \item $\sigma$: a parameter which controls how much of the noise in the data the kernel accounts for; naturally, 
    it moderates the risk of overfitting and underfitting.
    \item Length scale: a parameter that controls the influence - analogous to 
    the width each RBF neuron possesses - that each kernel (analogous to node) is to have.
\end{enumerate}

@@

A $4$-dimensional input space $X$ takes in an individual training observation $(\boldsymbol{\mathrm{x}}_{i}, y_i)$ and 
calculates the posterior weights distribution using \eqref{eq:nine}. 
Each possible weight vector $w$ represents a different function, explicitly 
based on the covariance function, that could fit the data. The expectation is then taken, which
effectively combines all weights and finds the mean, weighted by the 
respective probabilities. This represents the best functional estimate of the price movement. Based on this estimate 
the predictive output will be generated.

The predictive distribution for the test observations are derived similarly. The 
ultimate training process posterior weight distribution is used to derive the 
first test weight disttribution, which then averages over all $w$ again to find the
best functional estimate. It only requires past x and y.

\begin{align}
    p(y_{*}|x_{*}, X, y) \, dw = \mathcal{N}(\sigma_{n}^{2} x_{*}^{T}A^{-1}Xy, )
\end{align}

@@

The model returned a mean absolute error of MAE $=0.16$ and 
$R^2 = 9\%$ for the long-term prediction, outperforming the ARDL-ECM model by 
$2.4\%$.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{long-term-monthly.png}
    \caption{Long-term predicted vs monthly real FTSE250 price changes}
    \label{fig:lmonthly}
\end{figure}

The interpretation of this is that, over the long-run (4-5 years), the variations in the 
Consumer Price Index, interest rate, USD/GBP exchange rate, and M3 money 
supply explain or account for $9\%$ of the fluctuations of the
monthly FTSE250 value. This relates closely to the findings of \citeboth{conover1999}, 
who used an econometric model to establish that $4\%$ of the variation in monthly UK stock returns can be 
explained by variations in the money supply. It also suggests a higher level of influence 
than the ARDL-ECM model provided, that only $R^2=3.4\%$ of the monthly FTSE250 variations 
are affected by the selected macroeconomic variables, demonstrating the ability of machine learning
methods to capture subtle patterns in the data unavailable to econometric models.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{long-term-daily.png}
    \caption{Long-term predicted vs daily real FTSE250 price changes}
    \label{fig:ldaily}
\end{figure}

>>>>

% Define a command to create a hyperlink with only the figure number
\newcommand{\figref}[1]{%
    \hyperref[fig:#1]{Figure \ref{fig:#1}}%
}

% Define a command to create a hyperlink with only the algorithm number
\newcommand{\algoref}[1]{%
    \hyperref[alg:#1]{Algorithm \ref{alg:#1}}%
}

@@

\begin{abstract}
    This study investigates the impact of the Consumer Price Index, Bank of England Base Rate, US Dollar (USD) and British Pound Sterling (GBP) exchange rate and the money supply (M3) measure on the FTSE 250 index, and assesses the effectiveness of radial basis function neural networks as a tool to forecast the price of the FTSE 250 index based on the same set of macroeconomic indicators. This study shows quantitative estimates for the long-term and short-term impacts of the aforementioned macroeconomic indicators on the price of the FTSE 250 index, and demonstrates the effectiveness of radial basis function neural networks as a short-term forecasting tool for the price of the FTSE 250 index, thus providing an alternative to econometric forecasting models.

@@

The stock market has played a central role in the economic and 
political ascent of many nations ever since its inception \citep{levine2002}.  
However, whenever stock market instabilities have arisen in history they 
have had far-reaching destabilising effects. The Mississippi Bubble was 
an early example of the inherent dangers within stock markets in the form of 
speculative bubbles \citep{mississippibubble}. The role of
the macroeconomy in this crash was central. The French central bank, the 
\textit{Banque Royale}, aimed to convert its vast national debt into shares through the issuance of large 
amounts of government bonds and paper notes. This considerably increased the money supply, in turn 
greatly facilitating the ease of obtaining credit by individuals, many of whom invested in the Mississippi Company.
A mass sell-off began after the \textit{Banque Royale} announced the number of paper notes exceeded the 
value of metal coinage it possessed. This large-scale expansion of the money supply led to 
inflation burgeoning, reaching $23\%$ at its peak, and devastating France \citep{moen2001john}.  Nonetheless, stock markets also provide essential benefits by allowing companies to 
raise capital, offering investment opportunities for individuals, and 
supporting economic growth through efficient capital allocation. 
They enhance corporate governance through transparency and accountability, 
facilitate price discovery and market liquidity, and contribute to government 
financing through taxes \citep{shahbaz2016}. Effective and sustainable
macroeconomic policy can greatly aid in this beneficial process. Therefore, it is essential 
for economic policymakers to understand the impact of macroeconomic changes 
on the stock market.

The sensitivity of the stock market to macroeconomic indicators is also of great 
relevance to investors. Macroeconomic indicators provide 
insights into the overall health and direction of the 
economy, which affect investor sentiment and in turn influence market trends.
Changes in the macroeconomy have an effect on both consumers and firms 
which impacts their profitability, because these signals 
help investors anticipate future economic conditions and corporate profits
and consequently investment returns. Positive indicators tend to boost stock prices, while 
negative indicators can lead to declines in stock prices.

@@

Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability \citep{boe199section}. 
The empirical research supports this theory. \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003.
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange.
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the FTSE100. 
In addition, the flight-to-quality phenomenon that higher interest rates also encourage a shift from stocks to fixed-income 
assets such as bonds. \citeboth{asgharian2016} provided evidence for this in developed stock markets, including the UK. This suggests interest rates 
have a negative impact the stock market.

@@

One of the foundational studies on the relationship between the money supply and 
stock market prices was \citeboth{palmer1970}, who analysed the impact of M1 money supply on the S$\&$P 425 Industrial Stock Index. 
Palmer posited that trends in the nation's money stock could lead the 
private sector to adjust wealth portfolios, resulting in 
predictable movements in stock prices. \citeboth{homa1971} further argued that 
stock prices are significantly impacted by the risk-free interest rate, 
which is a function of money supply. This relationship was disputed by \citeboth{pesando1974}, 
who critiqued the theoretical and empirical methodology used in their study. In addition, \citeboth{sorensen1982} and 
\citeboth{bernanke2005} argue that only the unanticipated component of money supply changes 
affects the stock market price. Later research, however, such as \citeboth{bahloul2017}, 
\citeboth{synek2024}, and \citeboth{pícha2017} has found that the price of developed stock markets, such as the US and UK's, is in fact significantly influenced by the money supply. 
The nature of this relationship is disputed and varies across countries. \citeboth{homa1971} established a positive relationship between the 
stock market and the money supply. This was disputed by \citeboth{sellin2001}, who presents a negative theoretical relationship betwen the stock market and money supply. \citeboth{olawale2014} empirically found a significant positive impact by the money supply on stock returns in the US, but a negative one in the case of the UK.

@@

During the data collection process, it was necessary to deal with an issue raised by the fact that the Bank of England’s Monetary Policy Committee (MPC) 
does not convene on a monthly basis, and are not obligated to 
announce a rate on a regular basis. To address the resulting irregularities the data was adjusted using \algoref{interest_rate_adjustment}.

@@

The CPI, with an average of 88.36, has 
experienced significant fluctuations, highlighting periods of high inflationary pressure such as the onset of the Russia-Ukraine conflict in 2022, and periods of stable, low
inflation such as the early to late 2010s.

@@

Prior to constructing the ARDL model a natural logarithmic transform was applied
on the variables. This was done so that the coefficients appear as elasticities, 
which we can interpret easily as percentage changes. Subsequently,
the level at which the variables are stationary was ascertained using the 
Augmented Dickey-Fuller (ADF) unit root test. The optimum lag for the variables in the ARDL model was then determined using the Schwarz
Information Criterion.

The implicit form \eqref{eq:implicit} was represented in the ARDL model as:

@@

Here, the intercept term was given by $\alpha_0$. The 
remaining $\alpha_i$ terms represent the short-run relationships 
between the variables, and $\phi_j$ represent the long-run relationships.

@@

for, $i,j = 1,2,3,4,5$, with $i\neq j$. The F-statistic was computed for the joint
significance of the lagged variables and compared against the critical values
given by \citeboth{pesaran2001} to determine whether cointegration exists, 
and at what significance level. If cointegration exists, then 
the long-run coefficients $\kappa_i$ are estimated to find the cointegration equation:

@@

Then, an Error Correction Model (ECM) was constructed:

@@

where the error correction term, ECT, was found by calculating the residual 
produced by \eqref{eq:two}:

@@

The parameter $\gamma$ describes the speed at which short-run deviations from the long-run equilibrium
are adjusted. A negative $\gamma$ implies that there exists a correction mechanism that converges the deviation 
to the long-run equilibrium. To estimate the short-run coefficients, $\alpha_i$ and $\beta_k$, as 
well as the short-run correction speed $\gamma$ Ordinary Least Squares (OLS) regression was used.

Finally, diagnostic tests are conducted to confirm that the model does not
display autocorrelation, heteroskedascity, non-normality and that the cumulative 
sum of residuals and squared residuals are both stable.

@@

The null hypothesis of the ADF Unit Root test is that of non-stationarity, and the alternate hypothesis 
is that the variables are stationary.

@@

Subsequently, the optimum lag of the variables was calculated using the 
Schwarz Information Criterion, to obtain the optimum model 
of ARDL(5, 4, 7, 7, 4). The resulting F-statistic of $4.1$ was compared against the critical values
given by \citeboth{pesaran2001}, providing evidence of cointegration 
at the $10\%$, $5\%$ and $2.5\%$ significance levels, but not at the $1\%$ 
level.

@@

The long-run and short-run coefficients were then estimated, and presented 
alonsgide the diagnostic test results which confirm that the model does not
display autocorrelation, heteroskedascity, non-normality and the cumulative 
sum of residuals and squared residuals are both stable. The error
correction term is also given, and it was negative, as expected.

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & $-$ & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{***}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{****}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.1794^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

@@

Unsurprisingly, the USD/GBP exchange rate was significantly related to the FTSE250's movements. It 
should not be forgotten that, although it is considered to be a more domestic indicator of the LSE, 
at times $50\%$ of the FTSE250's earnings have come from abroad, indicating that 
the value of the Pound has a significant impact on this stock market. 
In the long run, the appreciation of the Pound would increase the relative 
cost of UK exports increases, thereby reducing the 
competitiveness of FTSE250 companies in international markets. This in turn would lower their earnings
and depress the FTSE250. Moreover, foreign investors may view the stronger Pound as a signal of reduced returns on their investments in the UK, which can lead to decreased investment in the FTSE250. 
In the short-term, however, before structural shifts can materialise we see that the
weakening of the Pound has a positive effect. As the USD has been 
the global reserve currency for many decades, the depreciation of the Pound 
against it will lead to increased revenue in Pound terms for FTSE250 companies with substantial overseas operations. This is because their foreign earnings, when converted back into Pounds, will be worth more, potentially increasing profits. The findings 
are in line with \citeboth{neifar2023} and \citeboth{khan2018}.

@@

The error correction term 
was negative and statistically significant ensuring that long-run equilibrium 
can be achieved. The adjustment speed of $0.1794$ indicates that 
roughly $18\%$ of the fluctuation in the previous month's price caused by the 
macroeconomic variable movements is adjusted back to the long-run equilibrium in the subsequent month. This implies 
that it takes approximately 6 months for the FTSE250 to return to its long-run equilibrium following 
changes in the selected macroeconomic variables.

@@

\begin{figure}[h]
    \centering
    \begin{tikzpicture}[shorten >=1pt,->,draw=black!50, node distance=2cm']

        % Labels for the layers
        \node[align=center] (inputlabel) at (-2,-2.8) {Input Layer};
        \node[align=center] (hiddenlabel) at (3,-7) {Hidden Layer};
        \node[align=center] (hiddenlabel) at (4,-0.3) {Bias};
        \node[align=center] (outputlabel) at (8,-2.8) {Output Layer};

        % Input Layer Nodes
        \node[circle, draw, minimum size=1cm] (I-1) at (0,-1) {$x_1$};
        \node[circle, draw, minimum size=1cm] (I-2) at (0,-2.5) {$x_2$};
        \node[] (I-dots) at (0,-4) {$\vdots$};
        \node[circle, draw, minimum size=1cm] (I-d) at (0,-5.5) {$x_d$};

        % Hidden Layer Nodes
        \node[circle, draw, minimum size=1cm] (H-0) at (3,-0.5) {$\phi_0$};
        \node[circle, draw, minimum size=1cm] (H-1) at (3,-2) {$\phi_1$};
        \node[] (H-dots) at (3,-3.5) {$\vdots$};
        \node[circle, draw, minimum size=1cm] (H-M) at (3,-6) {$\phi_M$};

        % Output Layer Node
        \node[circle, draw, minimum size=1cm] (O) at (6,-2.75) {$y$};

        % Connect input layer to hidden layer (excluding phi_0)
        \foreach \i in {1,2,d}
            \foreach \j in {1,M}
                \draw[->] (I-\i) -- (H-\j);

        % Connect phi_0 directly to output layer
        \draw[->] (H-0) -- (O);

        % Connect other hidden layer nodes to output layer
        \foreach \j in {1,M}
            \draw[->] (H-\j) -- (O);

    \end{tikzpicture}
    \caption{RBF-NN Architecture}
    \label{fig:rbf_nn_architecture}
\end{figure}

@@

The radial basis function approach 
introduces a set of $N$ basis functions, one for each 
data point. The radial basis function, $\phi(\cdot)$, is some non-linear function of the Euclidean distance
$||\boldsymbol{\mathrm{x}} - \boldsymbol{\mathrm{x_n}}||$. The output of the mapping is then expressed as a linear combination of these basis functions:

@@

By introducing several modifications to the exact interpolation 
procedure, the RBF-NN
model is obtained, as presented in \figref{rbf_nn_architecture}. 
The RBF-NN mapping takes the following form:

@@

where $\boldsymbol{c_j}$ is a RBF centre and $f$ is the radial basis function \citep{bishop1995}.

@@

The RBF-NN was utilised for both short-term and long-term 
forecasting. For the short-term, 
the model was tested over a 5-month horizon, based on the optimum lag calculated earlier. 
The long-term forecst spanned 57 months. 
The data was differenced based on the recommendations of the 
unit root tests conducted earlier to achieve stationarity.

The data was normalised using a z-score normalisation, because neural networks work best with data
that is between a specified range \citep{mendelsohn1993}. Following this 
it was necessary to program the RBF-NN into Python. The kernel regression approach outlined by 
\citeboth{bishop1995}, using the GaussianProcessRegressor package in Python was employed to model the RBF-NN.

@@

The GaussianProcessRegressor function in Python's Scikit library is based on the 
standard function

@@

where $y$ differs from the function values $f(x)$ by noise which is assumed
to follow an iid Gaussian distribution
$\mathcal{N}_1$
\begin{align*}
    \varepsilon \sim \mathcal{N}_1 \bigl(0, \sigma_{n}^2\bigr)
\end{align*}
This assumption together with the model produces the 
probability density of the observations given the inputs, 
which is factored over cases in the training set. This is
permissible because of the independence assumption \citep{rasmussen2006}.

@@

The weights \( w \) represent the parameters that define the relationship between the inputs and outputs. 
They are treated as random variables defined by the Gaussian distribution \eqref{eq: weight}.

A covariance function or kernel
is specified according to the characteristics of the data. In our model this 
is the generalised inverse multiquadric

\begin{align}
    k(x,x') = \biggl( ||x-x'|| + \sigma^2\biggr)^{-\beta} \label{eq: gimq}
\end{align}
and used to construct the covariance matrix $K$, which along with 
a mean function $m(x)$ forms the distribution of the the weights

\begin{align}
    w \sim \mathcal{N}\bigl(m(x), K\bigr) \label{eq: weight}
\end{align}

The mean function $m(x)$ acts as a bias, 
providing a baseline expectation of the function values. 
This forms the prior distribution over the weights \eqref{eq: weight}.

Then, by Bayes' Theorem,

\begin{align}
    p(w|y, X) = \frac{p(y|X,w) p(w)}{p(y|X)} \label{eq: post}
\end{align}
Using this, the posterior distribution \eqref{eq:nine} is obtained \citep{rasmussen2006}.

\begin{align}
    p(w|y, X) \sim \mathcal{N}\bigl(\frac{A^{-1}Xy}{\sigma_{n}^{2}}, A^{-1}\bigr) \label{eq:nine}
\end{align}

The covariance matrix $A^{-1} = K^{-1} + \sigma_{n}^{-2}(XX^T$).
All possible $w$, weighted by their probability, are then averaged to obtain the
most likely functional estimate of the output \citep{rasmussen2006}.

@@

First, the optimal hyperparameters in the covariance function \eqref{eq: gimq} 
are computed using GridSearch Cross Validation; these hyperparameters are:
\begin{enumerate}
    \item $\alpha$: a regularisation parameter representing a small quantity
    that is added to the diagonals of the covariance matrix to ensure it is 
    invertible. 
    \item $\beta$: the power in the covariance function, which controls the shape of the function.
    \item $\sigma$: a parameter which controls how much of the noise in the data the kernel accounts for. It moderates the risk of overfitting and underfitting.
    \item Length scale: a parameter that controls the influence that each kernel is to have. Analogous to 
    the width each RBF neuron possesses.
\end{enumerate}

@@

A $4$-dimensional input space $X$ takes in an individual training observation $(\boldsymbol{\mathrm{x}}_{i}, y_i)$ and 
calculates the posterior weights distribution. 
Each possible weight vector $w$ represents a different function, explicitly 
based on the RBF covariance function, that could fit the data. In the 
hidden layer the expectation is then taken, which
combines all weights and finds the mean, weighted by the 
respective probabilities. This represents the best functional estimate of the price movement. Based on this estimate 
the predicted output generated.

The testing process is conducted similarly. The 
posterior weight distribution that contains the information of the training phase becomes the new 
prior distribution. From this the posterior weight distributions for the test observations \eqref{eq: testw}
are calculated, and then the predictive output is found \citep{rasmussen2006}.

\begin{align}
    p(y_{*}|x_{*}, X, y) \, dw = \mathcal{N}(\sigma_{n}^{2} x_{*}^{T}A^{-1}Xy, x_{*}^{T}A^{-1}x_{*}) \label{eq: testw}
\end{align}

@@

The model returned a mean absolute error of MAE $=0.16$ and 
$R^2 = 9\%$ for the long-term prediction, providing a 
$2.4\%$ increase in the $R^2$ obtained by the ARDL-ECM model.

\begin{figure}[h!]
    \centering
    \input{long_term_monthly.tex}  
    \caption{Long-term predicted vs real monthly FTSE250 changes}
    \label{fig:longmonthly}
\end{figure}
The interpretation of this is that, over the long-run (4-5 years), the variations in the 
Consumer Price Index, interest rate, USD/GBP exchange rate, and M3 money 
supply explain or account for $9\%$ of the fluctuations of the
monthly FTSE250 value. This relates closely to the findings of \citeboth{conover1999}, 
who used an econometric model to establish that $4\%$ of the variation in monthly UK stock returns can be 
explained by variations in the money supply. It also suggests a higher level of influence 
than the ARDL-ECM model provided, that only $R^2=3.4\%$ of the monthly FTSE250 variations 
are affected by the selected macroeconomic variables, demonstrating the ability of machine learning
methods to capture subtle patterns in the data unavailable to econometric models.

\begin{figure}[h!]
    \centering
    \input{long_term_daily.tex}  
    \caption{Long-term predicted vs real daily FTSE250 changes}
    \label{fig:longdaily}
\end{figure}
